---
title: "Assignment #1 Literate Programming"
author: "Kyle Martin"
format: html
code-fold: true
toc: true
toc-depth: 3
toc-title: Table of Contents
number-sections: true
number-depth: 3
---
# The University of Idaho

BCB 520: Data Visualization Course!😁

## DATA

For this assignment we will be looking at how summary statistics can sometimes be misleading and highlight the need for data visualization in regards to understanding ones data set. 
To learn more about Quarto see <https://quarto.org>.

## Running Code

**Anscombe’s Quartet**
The four x-y pairs have identical summary statistics:

```{r}
#| output: false
rm(list = ls())
library(ggplot2)
library(grid)
library(gridExtra)
library(datasets)
library(tidyverse)
library(dplyr)
library(chron)
library(sp)
library(car)
library(lme4)
library(overlap)
library(AICcmodavg)
library(rainbow)
library(viridis)
```

```{r}
datasets::anscombe
```

## Example Problem

**Hypothesis for Replicates**

## Summary Statistics 

```{r}
tidy_anscombe <- anscombe %>%
 pivot_longer(cols = everything(),
              names_to = c(".value", "set"),
              names_pattern = "(.)(.)")
tidy_anscombe_summary <- tidy_anscombe %>%
  group_by(set) %>%
  summarise(across(.cols = everything(),
                   .fns = lst(min,max,median,mean,sd,var),
                   .names = "{col}_{fn}"))
#> `summarise()` ungrouping output (override with `.groups` argument)

vars<-c("set", "x_mean", "x_var",  "y_mean", "y_var")
Stuff<- as.data.frame(tidy_anscombe_summary[vars])
knitr::kable(Stuff)
```

## Visualization reveals hidden patterns! 

```{r}
ggplot(tidy_anscombe,
       aes(x = x,
           y = y)) +
  geom_point() +
  geom_point(data = tidy_anscombe_summary, aes(x=x_mean, y = y_mean, color = "blue", size = 4),
             show.legend = FALSE)+
  facet_wrap(~set) +
  geom_smooth(method = "lm", se = FALSE)
```